Anomaly detection with kernel density estimation on manifolds

### 2022-02-21 paper/paper.Rmd
Fix latex issue with n\_outlier

---
title: "Anomaly detection with kernel density estimation on manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
abstract: ""
# keywords: "manifold learning"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: true
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
---

@@

In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings. For a high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get a low-dimensional embedding. The manifold learning algorithms map the points through nonlinear functions that stretches some regions of the space while shrinks others. @Perrault2009 gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric and the Riemannian metric is a positive simi-definite square matrix for each data point. To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric. The outliers could then be defined as the points with lowest density estimates.

@@

<!-- $$ -->
<!-- \hat{f}(\mathbf{x})=\sum\limits_{i=1}^N K_{\mathbf{H}}(\mathbf{x}-\mathbf{x}_i) -->
<!-- $$ -->

@@

<!-- $$ -->
<!-- K_{\mathbf{H}}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}^{-1}(\mathbf{x}-\mathbf{x}_i)\right] -->
<!-- $$ -->

<!-- The matrix $\mathbf{H}$ is called the bandwidth matrix and is very important.  -->

@@

$$
\hat{f}(\mathbf{x})=\sum\limits_{i=1}^N K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)\\
K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}_i^{-1}(\mathbf{x}-\mathbf{x}_i)\right]
$$

@@

  \STATE Calculate the Riemannian metric $h(x)$ as the rank $d$ pseudo inverse of $\tilde{h}(x)$ with 
  $$
    h(x) = U diag{1/(\Gamma[1:d])} U^\prime,
  $$
  where $[U, \Gamma]$ is the eigendecomposition of matrix $\tilde{h}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Gamma$ in descending order.

@@

\begin{algorithm}[!htb]
  \caption{Variable kernel density estimates with Riemannian metric}
  \label{alg:vkderm}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $x_i$ for all $i=1,\ldots,N$ }
  \Output{ outliers embedding coordinates $y_1, \dots, y_{\textit{n_outliers}}$ with their estimated densities $f_1, \dots, f_{\textit{n_outliers}}$ }
  \Parameter{ number of outliers \textit{n_outliers}, embedding dimension $d$ }
  \OptParameter{ }
  \BlankLine
  \begin{algorithmic}[1]

  \STATE for all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $h_i$ using Algorithm&nbsp;\@ref(alg:learnmetric) with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;

  \STATE Set the variable bandwidth for each point as $h_i$;

  \STATE Compute the kernel density estimates for each point as
  $$
  \hat{f}(\mathbf{y})=\sum\limits_{i=1}^N K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)\\
  K_{\mathbf{H}_i}(\mathbf{x}-\mathbf{x}_i)=(2\pi)^{-d/2}|\mathbf{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\mathbf{x}-\mathbf{x}_i)'\mathbf{H}_i^{-1}(\mathbf{x}-\mathbf{x}_i)\right];
  $$

  \STATE Reorder the embedding coordinates $y$ according to the density estimates $f(y)$ and subset the top $\textit{n_outliers}}$ as the outliers.

@@

Now that we have proposed a way to take into account the distortion of manifold in kernel density estimate, it would be straightforward to produce the highest density region plots [HDR plots; @Hyndman1996-lk] which are also computed using kernel density estimate of the embedding. Similar to Figure \@ref(fig:hdr) in Chapter \ref{mlann}, we can use the HDR plots to do anomaly detection and apply it to very high dimensional data sets.

>>>>

---
title: "Anomaly detection with kernel density estimation on manifolds"
author:
- familyname: Cheng
  othernames: Fan
  address: Monash University
  email: Fan.Cheng@monash.edu
  # correspondingauthor: true
- familyname: Hyndman
  othernames: Rob J
  address: Monash University
  email: Rob.Hyndman@monash.edu
- familyname: Panagiotelis
  othernames: Anastasios
  address: University of Sydney
  email: Anastasios.Panagiotelis@sydney.edu.au
abstract: ""
keywords: "manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels"
# wpnumber: 03/2021
# jelcodes: C55, C65, C80
blind: true
cover: false
toc: false
lot: false
bibliography: references.bib
biblio-style: authoryear-comp
output:
  monash::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
      # after_body: appendix.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
    latex_engine: pdflatex
---

@@

In this section, we introduce the proposed method to detect anomalies based on the kernel density estimates of manifold learning embeddings where the Riemannian matrix is used as the pointwise variable bandwidth to measure the direction and angle of the distortion of the low-dimensional embeddings. For a high-dimensional data set, various manifold learning algorithms including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, are applied to get a low-dimensional embedding. The manifold learning algorithms map the points through nonlinear functions that stretches some regions of the space while shrinks others. @Perrault-Joncas2013-pq gives us an idea of how to measure the direction and angle of the distortion using the Riemannian metric and the Riemannian metric is a positive simi-definite square matrix for each data point. To learn the distribution of the low-dimensional embedding, we use the kernel density estimation with the bandwidth matrix being the Riemannian metric. The outliers could then be defined as the points with lowest density estimates.

@@

<!-- $$ -->
<!-- \hat{f}(\pmb{x})=\sum\limits_{i=1}^N K_{\pmb{H}}(\pmb{x}-\pmb{x}_i) -->
<!-- $$ -->

@@

<!-- $$ -->
<!-- K_{\pmb{H}}(\pmb{x}-\pmb{x}_i)=(2\pi)^{-d/2}|\pmb{H}|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{x}_i)'\pmb{H}^{-1}(\pmb{x}-\pmb{x}_i)\right] -->
<!-- $$ -->

<!-- The matrix $\pmb{H}$ is called the bandwidth matrix and is very important.  -->

@@

$$
\hat{f}(\pmb{x})=\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{x}_i)\\
K_{\pmb{H}_i}(\pmb{x}-\pmb{x}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{x}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{x}_i)\right]
$$

@@

  \STATE Calculate the Riemannian metric $h(x)$ as the rank $d$ pseudo inverse of $\tilde{h}(x)$ with 
  $$
    h(x) = U diag{1/(\Lambda[1:d])} U^\prime,
  $$
  where $[U, \Lambda]$ is the eigendecomposition of matrix $\tilde{h}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.

@@

\begin{algorithm}[!htb]
  \caption{Variable kernel density estimates with Riemannian metric}
  \label{alg:vkderm}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}
  \Input{ high-dimensional data $x_i$ for all $i=1,\ldots,N$ }
  \Output{ outliers embedding coordinates $y_1, \dots, y_{n\_outliers}$ with their estimated densities $f_1, \dots, f_{n\_outliers}$ }
  \Parameter{ number of outliers $n\_outliers$, embedding dimension $d$ }
  \BlankLine
  \begin{algorithmic}[1]

  \STATE For all $i=1,\ldots,N$, compute the $d$-dimensional embeddings $y_i$ with any exsiting manifold leanring algorithms and the corresponding Riemannian metric $h_i$ using the Learn metric algorithm with inputs $d$ and $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ for heat kernels;

  \STATE Set the variable bandwidth for each point as $\pmb{H}_i = h_i$;

  \STATE Compute the kernel density estimates for each point as
  $$
  \hat{f}(\pmb{y})=\sum\limits_{i=1}^N (2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{y}-\pmb{y}_i)'\pmb{H}_i^{-1}(\pmb{y}-\pmb{y}_i)\right];
  $$

  \STATE Reorder the embedding coordinates $y$ according to the density estimates $f(y)$ and subset the top \textit{n\_outliers} as the outliers.

@@

Now that we have proposed a way to take into account the distortion of manifold in kernel density estimate, it would be straightforward to produce the highest density region plots [HDR plots; @Hyndman1996-lk] which are also computed using kernel density estimate of the embedding. 
<!-- Similar to Figure \@ref(fig:hdr) in Chapter \@ref{mlann}, we can use the HDR plots to do anomaly detection and apply it to very high dimensional data sets. -->
